Depression and suicidality as credible signals of need

### 2021-05-25 Paper.rmd
Minor edits

Raymond Firth was one of the first to view suicidality as a gamble to improve one's circumstances in the here and now. Based on observations that suicide attempts often followed loss or conflict and varied substantially in their likelihood of death, he argued that a sizable subset of the suicide attempts among the Tikopia were not meant to end in death but instead were a means to elicit attention, status, or immediate reintegration into the community following negative events [@firth1936; @Firth:1961].

@@

In the bargaining framework, the willingness to delay credibly signals the valuation of current cooperative arrangements, with greater delays indicating lower valuations. The primary aim of this study was to examine responses to emotional signals that varied in the extent to which they reduced productivity, which we refer to as costs, in an experimental vignette study in which a possible victim of adversity asks for help from the participant, but has incentives to exaggerate her need. As signal cost increased, we predicted that participants would report (1) increased belief in the signaler's claims and (2) increased likelihood of providing help, with (3) the increased likelihood of providing help mediated by the increased belief in the signaler's need.

@@

Participants for this study were recruited from Amazon Mechanical Turk, a crowdsourcing platform that allows for the creation of Human Intelligence Tasks (HITs) that workers can complete for pay. As Amazon provides the infrastructure, it allows for a relatively low-cost way of collecting data for academic research at the cost of the data not being representative of any real population [@thomas2017]. Despite this limitation, MTurk samples have a wider range of ages and incomes than most university samples [@kennedy2020; @DworkinEtAl2016; @thomas2017]. US MTurk samples also differ from the general US population, mainly in being younger, more educated, and lower income [@RossEtAl2010; @boas2020].

@@

**Basketball coach vignette**: Each participant was asked to imagine that they are a university athletic director. The star player on the women's basketball team comes to the participant and claims she was sexually assaulted by her head coach, a physically powerful man. However, there is a history of conflict between the star player and the coach over playing time, and police are unable to find evidence to corroborate her claims.

**Brother-in-law vignette**: Each participant was asked to imagine that they let their sister, brother-in-law, and niece move in with them after their sister's family lost their house in a fire. During this time, the participant's 15 year old daughter becomes jealous of her cousin, who appears to be a social competitor. A few weeks after claiming her cousin was trying to steal her boyfriend, the participant's daughter accuses her uncle of sexually assaulting her.

**Romantic partner vignette**: Each participant was asked to imagine that they found a highly desirable romantic partner after years of being single. However, the participant's 13-year-old daughter, who has a history of interfering with the participant's past relationships, is clearly unhappy with the new partner. After a period of sustained conflict with both the participant and the romantic partner, she accuses the romantic partner of physically assaulting her, but cannot produce any evidence.

**Thwarted marriage vignette**: Indian sample only. Each participant was asked to imagine that their family was trying to arrange a dowry for their older daughter (the signaler in this vignette) so she can marry a man she already loves, while still saving enough money for their younger daughter's dowry. After the man's family demands more money, the participant's family tries to find a second man, who the older daughter claims to find unattractive. Any increase in the dowry will come at the younger daughter's expense. The participant therefore proceeds to arrange a marriage to the second man as the first man's family makes arrangements with a different woman.

@@

We also asked which emotions participants felt the signaler was experiencing using a multiple choice question in which they could select as many options as they would like. This included emotions directly related to the signals (e.g., sad, depressed, and suicidal), states which suggest genuine need (traumatized and violated), states which suggest deception (e.g., deviousness or jealousy), and was the victim mentally ill. The complete list can be found in the SI. To further explore the effect of signaling on participants' inferences of the signaler's emotional state, we created two new variables: *Low mood* was the sum of the binary variables *Depressed*, *Distressed* and *Sad*; and *Manipulative* was the sum of the binary variables *Devious* and *Jealous.*

@@

As both a validity check and a way to understand the degree participants would be willing to help if they believed the participant completely, we presented participants with strong evidence that the claims were true. In the US sample, this involved telling participants there was video evidence of the event in question occurring or a similar event after the fact. With the thwarted marriage vignette, this involved the participant seeing the man their daughter wants to marry trash talk their daughter and their family. Participants were then asked to rate their likelihood of acting (*T3 Action*).

@@

We therefore fit models of our two outcome variables (which we rescaled to the continuous interval [0, 1]) in five ways. First, we fit the preregistered OLS models (reported in the SI). Second, we fit fractional regression models, introduced in a seminal paper by @papke1996 for continuous outcomes on the interval [0, 1], which, in brief, involve fitting a quasi-likelihood version of logistic regression with robust standard errors. This method is implemented in the `frm` package [@ramalho2019]. Third, we fit GLM logistic regression models, computing 95% CI's via bootstrapping, using the `glmmboot` package created to model continuous outcomes on [0, 1] [@humphrey2020]. Fourth, we fit GLM models with the quasibinomial family and logit link using the `glm` function in R stats library, which can fit continuous outcomes on [0, 1]. Finally, we computed robust standard errors and 95% confidence intervals from the quasibinomial models using the `sandwich` package [@zeileis2006]. The estimated coefficients, standard errors, and confidence intervals produced by the latter four methods were *very* similar, and often identical (see Figure \@ref(fig:modelcompareplots). We therefore proceeded to fit and report models using `glm` with the quasibinomial family and logit link because model objects produced by this function are compatible with numerous other R packages for summarizing and plotting effects.

@@

To test our preregistered hypothesis that the likelihood of acting to help the victim would be largely mediated by a signal's positive impact on the participant's belief in the victim's need, we used the `mediation` package [@tingley2014] to fit a mediation model for *Depression* treatment vs. *Verbal request* control. See Figure \@ref(fig:medmodel).

@@

Across the four vignettes, mean belief of the victim was relatively low at baseline (T1), Mean = `r signif(mean(d$T1Belief, na.rm = T), 2)`, albeit with wide variation, SD = `r signif(sd(d$T1Belief, na.rm=T), 2)`; `r sum(d$T1Belief==0)` participants (`r 100*signif(sum(d$T1Belief==0)/nrow(d), 2)`%) rated their belief = 0, and `r sum(d$T1Belief==1)` participants (`r 100*signif(sum(d$T1Belief==1)/nrow(d), 2)`%) rated it as = 1. The distribution of likelihood of helping the victim (action) was similar, Mean = `r signif(mean(d$T1Action, na.rm = T), 2)`, SD = `r signif(sd(d$T1Action, na.rm=T), 2)`, with `r sum(d$T1Action==0)` participants (`r 100*signif(sum(d$T1Action==0)/nrow(d), 2)`%) rating their action = 0, and `r sum(d$T1Action==1)` participants (`r 100*signif(sum(d$T1Action==1)/nrow(d), 2)`%) rated action as = 1. Although not a pre-registered prediction, *T1 Belief* and *T1 Action* were highly correlated across the four vignettes, consistent with help being worth providing if the signaler's claims were true (see Figure \@ref(fig:t1plot)).

@@

The effects of the signals on rated beliefs and actions differed substantially by vignette, which we display in three ways for each signal in each vignette: estimated cumulative distributions plots, which show the entire distributions of rated beliefs and actions, including which conditions have high fractions of 0's and 1's (Figures \@ref(fig:ecdf)); regression estimates of the mean effects of the signals on beliefs and actions by vignette (Figure \@ref(fig:beliefvignette)); and between- and within-subjects signal effect sizes by vignette (Figures \@ref(fig:effectsizes) and \@ref(fig:effectsizes2)).

@@

The effect of *Suicide attempt* on belief at T2 was similar to the *Depression* conditions across vignettes, but it resulted in greater likelihood of support. One interpretation is that although some participants did not believe the victim's story, her signal nevertheless convinced them that she needed help. For example, maybe the brother-in-law did not assault her, but the presence of his family in her home was causing genuine distress. Support for this interpretation comes from our mediation analyses, which showed that the likelihood of help was largely, but not entirely, mediated by signal's effect on belief in need.

There were minor associations of age and sex with belief and action in the US participants, with both being higher among females and younger individuals. The US vignettes all involved assaults against young women, which might have been more salient to female and younger participants. In the Indian sample, high baseline belief and action were somewhat lower among those with more education and among females, respectively. Costlier signals, suicidality in particular, had a larger effect among older individuals, perhaps because older individuals were more likely to have children of marriageable age, like the victim in the vignette.

@@

Depression is costly, sometimes resulting in death by suicide. Our results suggest that these costs, which mainstream theories take as evidence of brain dysfunction [@SymeHagen2021], function to help victims of adversity elicit support when their true level of need is private information and they have conflicts with social partners. These findings align with real-world evidence that depression simultaneously elicits positive and negative responses from social partners [@SymeHagen2021]. If the bargaining model is correct, depression is an important social problem, not a medical one [@syme2020].

@@

```{r division, fig.width=9, fig.cap="Effect of signals on division of dowry between the daughters in the thwarted marriage vignette, with an interaction between T1 Division and signal. Means estimated with a generalized linear regression model with the quasibinomial family."}
effectsPlots_India$m7plot
```

```{r divisionage, fig.width=9, fig.cap="Interaction of age with signals on *T2 Division* of dowry between the daughters in the thwarted marriage vignette."}
effectsPlots_India$m8plot
```

>>>>

Raymond Firth, an anthropologist who worked in the southwestern Pacific, was one of the first to view suicidality as a gamble to improve one's circumstances in the here and now. Based on observations that suicide attempts often followed loss or conflict and varied substantially in their likelihood of death, he argued that a sizable subset of the suicide attempts among the Tikopia were not meant to end in death but instead were a means to elicit attention, status, or immediate reintegration into the community following negative events [@firth1936; @Firth:1961].

@@

In the bargaining framework, the willingness to delay credibly signals the valuation of current cooperative arrangements, with greater delays indicating lower valuations. The primary aim of this study was to examine responses to emotional signals that varied in the extent to which they reduced productivity or put future productivity at risk, which we refer to as costs, in an experimental vignette study in which a possible victim of adversity asks for help from the participant, but has incentives to exaggerate her need. As signal cost increased, we predicted that participants would report (1) increased belief in the signaler's claims and (2) increased likelihood of providing help, with (3) the increased likelihood of providing help mediated by the increased belief in the signaler's need.

@@

Participants for this study were recruited from Amazon Mechanical Turk, a crowdsourcing platform that allows for the creation of Human Intelligence Tasks (HITs) that workers can complete for pay. As Amazon provides the infrastructure, it allows for a relatively low-cost way of collecting data for academic research, with the disadvantage that the data are not representative of any real population [@thomas2017]. Despite this limitation, MTurk samples have a wider range of ages and incomes than most university samples, and therefore might be more informative about the general population [@kennedy2020; @DworkinEtAl2016; @thomas2017]. US MTurk samples do differ from the general US population, though, mainly in being younger, more educated, and lower income [@RossEtAl2010; @boas2020].

@@

**Basketball coach vignette**: Participants were asked to imagine that they are a university athletic director. The star player on the women's basketball team comes to the participant and claims she was sexually assaulted by her head coach, a physically powerful man. However, there is a history of conflict between the star player and the coach over playing time, and police are unable to find evidence to corroborate her claims.

**Brother-in-law vignette**: Participants were asked to imagine that they let their sister, brother-in-law, and niece move in with them after their sister's family lost their house in a fire. During this time, the participant's 15 year old daughter becomes jealous of her cousin, who appears to be a social competitor. A few weeks after claiming her cousin was trying to steal her boyfriend, the participant's daughter accuses her uncle of sexually assaulting her.

**Romantic partner vignette**: Participants were asked to imagine that they found a highly desirable romantic partner after years of being single. However, the participant's 13-year-old daughter, who has a history of interfering with the participant's past relationships, is clearly unhappy with the new partner. After a period of sustained conflict with both the participant and the romantic partner, the daughter accuses the romantic partner of physically assaulting her, but cannot produce any evidence.

**Thwarted marriage vignette**: Indian sample only. Participants were asked to imagine that their family was trying to arrange a dowry for their older daughter (the signaler in this vignette) so she can marry a man she already loves, while still saving enough money for their younger daughter's dowry. After the man's family demands more money, the participant's family tries to find a second man, who the older daughter claims to find unattractive. Any increase in the dowry will come at the younger daughter's expense. The participant therefore proceeds to arrange a marriage to the second man as the first man's family makes arrangements with a different woman.

@@

We also asked which emotions participants felt the signaler was experiencing using a multiple choice question in which they could select as many options as they would like. This included emotions directly related to the signals (e.g., sad, depressed, and suicidal), states which suggest genuine need (traumatized and violated), states which suggest deception (e.g., deviousness or jealousy), and if the victim mentally ill. The complete list can be found in the SI. To further explore the effect of signaling on participants' inferences of the signaler's emotional state, we created two new variables: *Low mood* was the sum of the binary variables *Depressed*, *Distressed* and *Sad*; and *Manipulative* was the sum of the binary variables *Devious* and *Jealous.*

@@

As both a validity check and a way to understand the degree participants would be willing to help if they believed the participant completely, at T3 we presented participants with strong evidence that the claims were true. In the US sample, this involved telling participants there was video evidence of the event in question occurring or a similar event after the fact. With the thwarted marriage vignette, this involved the participant seeing the man their daughter wants to marry trash-talk their daughter and their family. Participants were then asked to rate their likelihood of acting (*T3 Action*).

@@

We therefore fit models of our two outcome variables with the preregistered OLS models (reported in the SI) and four variants of logistic regression (for which we rescaled the 0-100 outcomes to the continuous interval [0, 1]). First, we fit fractional regression models, introduced in a seminal paper by @papke1996 for continuous outcomes on the interval [0, 1], which, in brief, involve fitting a quasi-likelihood version of logistic regression with robust standard errors. This method is implemented in the `frm` package [@ramalho2019]. Second, we fit GLM logistic regression models, computing 95% CI's via bootstrapping, using the `glmmboot` package created to model continuous outcomes on [0, 1] [@humphrey2020]. Third, we fit GLM models with the quasibinomial family and logit link using the `glm` function in R stats library, which can fit continuous outcomes on [0, 1]. Fourth, we computed robust standard errors and 95% confidence intervals from the quasibinomial models using the `sandwich` package [@zeileis2006]. The estimated coefficients, standard errors, and confidence intervals produced by these four methods were *very* similar, and often identical (see Figure \@ref(fig:modelcompareplots). We therefore proceeded to fit and report models using `glm` with the quasibinomial family and logit link because model objects produced by this function are compatible with numerous other R packages for summarizing and plotting effects.

@@

To test our preregistered hypothesis that the likelihood of acting to help the victim would be largely mediated by a signal's positive impact on the participant's belief in the victim's need, we used the `mediation` package [@tingley2014] to fit a mediation model for *Depression* treatment vs. *Verbal request* control. We did the same for *Suicide attempt*. See Figure \@ref(fig:medmodel).

@@

Across the four vignettes, mean belief of the victim (after rescaling original 0-100 values to 0-1) was relatively low at baseline (T1), Mean = `r signif(mean(d$T1Belief, na.rm = T), 2)`, albeit with wide variation, SD = `r signif(sd(d$T1Belief, na.rm=T), 2)`; `r sum(d$T1Belief==0)` participants (`r 100*signif(sum(d$T1Belief==0)/nrow(d), 2)`%) rated their belief = 0, and `r sum(d$T1Belief==1)` participants (`r 100*signif(sum(d$T1Belief==1)/nrow(d), 2)`%) rated it as = 1. The distribution of likelihood of helping the victim (action) was similar, Mean = `r signif(mean(d$T1Action, na.rm = T), 2)`, SD = `r signif(sd(d$T1Action, na.rm=T), 2)`, with `r sum(d$T1Action==0)` participants (`r 100*signif(sum(d$T1Action==0)/nrow(d), 2)`%) rating their action = 0, and `r sum(d$T1Action==1)` participants (`r 100*signif(sum(d$T1Action==1)/nrow(d), 2)`%) rated action as = 1. Although not a pre-registered prediction, *T1 Belief* and *T1 Action* were highly correlated across the four vignettes, consistent with help being worth providing if the signaler's claims were true (see Figure \@ref(fig:t1plot)).

@@

The effects of the signals on rated beliefs and actions differed substantially by vignette, which we display in three ways for each signal in each vignette: estimated cumulative distributions plots, which show the entire distributions of rated beliefs and actions, including which conditions have high fractions of 0's and 1's (Figures \@ref(fig:ecdf)); regression estimates of the mean effects of the signals on beliefs and actions by vignette (Figure \@ref(fig:beliefvignette)); and between- and within-subjects signal effect sizes (Cohen's d) by vignette (Figures \@ref(fig:effectsizes) and \@ref(fig:effectsizes2)).

@@

The effect of *Suicide attempt* on *T2 Belief* was similar to the *Depression* conditions across vignettes, but it resulted in greater *T2 Action*. One interpretation is that although some participants did not believe the victim's story, her signal nevertheless convinced them that she needed help. For example, maybe the brother-in-law did not assault her, but the presence of his family in her home was causing genuine distress. Support for this interpretation comes from our mediation analyses, which showed that the likelihood of help was largely, but not entirely, mediated by signal's effect on belief in need.

There were minor associations of age and sex with *T2 Belief* and *T2 Action* in the US participants, with both being higher among females and younger individuals. The US vignettes all involved assaults against young women, which might have been more salient to female and younger participants. In the Indian sample, *T2 Belief* and *T2 Action* were somewhat lower among those with more education and among females, respectively. Costlier signals, suicidality in particular, had a larger effect among older individuals, perhaps because older individuals were more likely to have children of marriageable age, like the victim in the vignette.

@@

Depression is costly, sometimes resulting in death by suicide. Our results suggest that these costs, which mainstream theories take as evidence of brain dysfunction [@SymeHagen2021], instead function to help victims of adversity elicit support when their true level of need is private information and they have conflicts with social partners. These findings align with real-world evidence that depression simultaneously elicits positive and negative responses from social partners [@SymeHagen2021]. If the bargaining model is correct, depression is an important social problem, not a medical one [@syme2020].

@@

```{r division, fig.width=9, fig.cap="Effect of signals on division of dowry between the daughters in the thwarted marriage vignette, with an interaction between T1 Division and signal. Division > 0.50 favored older daughter. Means estimated with a generalized linear regression model with the quasibinomial family."}
effectsPlots_India$m7plot
```

```{r divisionage, fig.width=9, fig.cap="Interaction of age with signals on *T2 Division* of dowry between the daughters in the thwarted marriage vignette. Division > 0.50 favored older daughter. "}
effectsPlots_India$m8plot
```
